State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-05-09 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
EDITS TO DISCUSSION. ADDED DESCRIPTIONS OF PARAGRAPHS

\section{Discussion}\label{DiscussionSection}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of neural fields. The level of detail of the estimated field is related to the basis functions decomposition, thereby providing a flexible approach where a trade-off can be made between physiological detail and computational complexity. The choice of Gaussian basis functions can be justified by the existence of the so called bump solutions for this class of model~\cite{Coombes2005}.

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. These tools can be also used to better design electrophysiological recording techniques to advance model-based neural parameter estimation in order to avoid spatial aliasing. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field field is spatially band-limited. The frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel, therefore some information will be lost due to the field decomposition. The implication of this loss of information is the high spatial frequency content of the connectivity kernel. This is demonstrated in Figure~\ref{fig:KernelEstimates} of the results where the frequency content of the estimated connectivity kernel is less than the true kernel.

Even though the high frequency content of the connectivity kernel is lost though the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters falls into a tight distribution as seen in Figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

To the authors best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. The importance of this can be illustrated by considering the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion or water in grey matter, the fine spatial scale and the high degree of crossing of fibres~\cite{Assaf2008}. In addition, other non-model based techniques for estimating functional connectivity such as Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999} are restricted to the spatial scales of the recording electrodes. The model based technique proposed in this paper estimates the connectivity structure independently of the recording electrode geometry. An advantage of the auto-regressive (AR) based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. The AR and the proposed model-based methods are complementary, where they estimating connectivity at different spatial scales. Therefore, by combining these methods a more realistic model can be estimated. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs where a Turing-Hoft bifurcation occurs~\cite{Hutt2005}, and types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to various oscillatory states. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

Parameter estimation may have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.

This estimation framework also has theoretical implications. This estimation framework has the potential to allow testing hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so called bump solutions are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition exploration of parameters and the state-space can be explored during other cognitive tasks and states of arousal such as sleep and anaesthesia.

In this paper the widths of the connectivity kernel basis functions are assumed to be known. In a future work we plan to extend this work to unknown widths. We hope to achieve this by using a sufficiently large set of connectivity basis functions within the physiological plausible limits, and letting estimation procedure take care of the weights. In addition, the assumption homogeneity can be relaxed, where heterogeneity can be modelled by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria.

@@

\section{Discrete Time Model}\label{Time Discretization} To form the IDE neural field model a time discretization must be perform. We used a one-step Euler method where Eq.~\ref{FinalFormContinuous} can be approximated by 

>>>>

\section{Discussion}\label{DiscussionSection}

\omg{NOTE: I HAVE ADDED A SENTENCE DESCRIBING SIMPLY WHAT I WANT TO SAY IN EACH PARAGRAPH. PLEASE USE THIS AS A GUIDE FROM EDITING.}

\omg{FIRST DESCRIBE AN OVER ALL SUMMARY OF WHAT WE HAVE DONE AND WHAT IS NOVEL}
This paper presents a novel framework for model reduction to facilitate practical state and parameter estimation of neural fields. The level of detail of the estimated field is related to the basis functions decomposition, thereby providing a flexible approach where a trade-off can be made between physiological detail and computational complexity. The choice of Gaussian basis functions can be justified by the existence of the so called bump solutions for this class of model~\cite{Coombes2005}.

\omg{DISCUSS RESULTS: FREQ ANALYSIS. USE THIS TO JUSTIFY KERNEL ERRORS.}
The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. These tools can be also used to better design electrophysiological recording techniques to advance model-based neural parameter estimation in order to avoid spatial aliasing. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel, therefore some information will be lost due to the field decomposition. The implication of this loss of information is the high spatial frequency content of the connectivity kernel. This is demonstrated in Figure~\ref{fig:KernelEstimates} of the results where the frequency content of the estimated connectivity kernel is less than the true kernel.

\omg{DISCUSS THAT EVEN THOUGH THE KERNEL ESTIMATE IS NOT EXACT, IT IS USEFUL TO KNOW THE RATIO OF EXCITATION TO INHIBITION AND THAT RELATIVE MEASURES CAN BE ACCURATELY TRACKED}
Even though the high frequency content of the connectivity kernel is lost though the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in Figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

\omg{NEED TO ADD SOMETHING ABOUT THE TIME CONSTANT}

\omg{DISCUSS THE IMPLICATION OF ESTIMATING INTRACORTICAL CONNECTIVITY AND WHERE WE FIT WITHIN OTHER FUNCTIONAL CONNECTIVITY RESEARCH AREAS}
To the authors best knowledge, this is the first paper to propose an estimation framework for intracortical connectivity from electrophysiological data. The importance of this can be illustrated by considering the connectivity structure at this scale can not be estimated using diffusion tensor imaging due to the near isotropic diffusion or water in grey matter~\cite{Assaf2008}. In addition, other non-model based techniques for estimating functional connectivity such as Granger causality \cite{Hesse2003}, the direct transfer function \cite{Kaminski1991}, and partial directed coherence \cite{Sameshima1999} are restricted to the spatial scales of the recording electrodes. The model-based technique proposed in this paper estimates the connectivity structure independently of the recording electrode geometry. An advantage of the auto-regressive (AR) based methods is that they allow for anisotropic connectivity, which is present in cortico-cortical fibres. It should be noted, however, that the AR and the proposed model-based methods are complementary, where connectivity is estimated at different spatial scales. Therefore, we envisage by combining these methods a more realistic model can be acquired. A recent theoretical study~\cite{Jirsa2009} has demonstrated the importance of this multi-scale structure in generating the characteristic rhythms of the brain.

\omg{DISCUSS A BIG PICTURE. WHERE THIS WORK SHOULD GO AND WHERE IT SHOULD BE APPLIED ON THE FUTURE}
Generating patient specific neural field models has the potential to have significant impact on several areas of neuroscientific research and clinical neurology. Specifically, the development of a patient-specific state-space model will allow for the application of systems theoretical techniques from engineering to be applied to neuro-dynamics. For example, state tracking could be used in brain-computer interface or epileptic seizure prediction applications, where specific regions of the state-space may indicate intent of movement or imminent seizures. Recent studies using neural field models have demonstrated the theoretical implications of specific connectivity structure of the cortical columns in neural field models, where the connectivity kernel governs where a Turing-Hoft bifurcation occurs~\cite{Hutt2005}, and types of oscillations that can be generated~\cite{Schmidt2009}. This implies that if we could estimate the connectivity structure for an individual, then we could capture patient specific neurodynamics that lead to hyper-synchronous oscillatory states such as Turing patterns. In addition, a patient specific state-space model will allow application of control engineering techniques to robustly prevent the neuro-dynamics entering pathological regions of state-space using electrical stimulation.

Parameter estimation may also have significant impact on the treatment of disease. For example, recent theoretical work has demonstrated that seemingly similar electrographic seizures in absence epilepsy may arise from abnormal excitatory or inhibitory mechanisms~\cite{Marten2009}. These very different mechanisms would require very different medications for successful treatment. However, the underlying mechanisms would be hidden from the clinician in the normal clinical setting. Successful parameter estimation has the potential to reveal these hidden mechanisms allowing for improved treatment strategies.

\omg{DISCUSS HOW IT CAN LINK THEORETICAL MODELLING STUDIES TO DATA.}
This estimation framework also has theoretical implications by establishing a framework that has the potential to allow testing hypotheses that have been generated in theoretical studies and to validate neural field models. For example, it has been hypothesised that so called bump solutions are a possible mechanism for short term memory formation~\cite{Coombes2005}. Using the proposed model-based framework, we can reconstruct a neural field from data, and estimate parameters and check if they correspond the the theoretically derived parameters where these bump solutions can exist. In addition exploration of parameters and the state-space can be explored during other cognitive tasks and states of arousal such as sleep and anaesthesia.

\omg{DISCUSS SHORT-COMINGS AND FUTURE WORK}
In this paper the widths of the connectivity kernel basis functions are assumed to be known. In a future work we plan to extend this work to unknown widths. We hope to achieve this by using a sufficiently large set of connectivity basis functions within the physiological plausible limits, and letting estimation procedure take care of the weights. In addition, the assumption homogeneity can be relaxed, where heterogeneity can be modelled by careful placement of field basis functions. For example, if cortical regions have a greater density of neural populations, then more basis functions can be used in these areas. In this scenario, model selection and basis function positioning can be achieved using instantaneous frequency analysis of the the neural field or alternative model selection tools based on information criteria.

@@

\section{Discrete Time Model}\label{Time Discretization} To form the IDE neural field model a time discretisation must be perform. We used a one-step Euler method where Eq.~\ref{FinalFormContinuous} can be approximated by 
